|

### 2025-10-31 manuscript/paperCMSACversion.Rmd
cmsac version of paper.

However, using mutual information in this form does not suit our needs
in this setting. The problem is that mutual information in this form
will yield high values of mutual information when the output from the
tournament $\hat{r}$ is highly consistent *even if the ranking from the tournament is incorrect* (*i.e.* the rankings produced by the tournament are very different from the true rankings). As an example, if $n=4$ and the true order of
$\boldsymbol{\theta}$ is $r = \{1, 2, 3, 4\}$ and
$p(\hat{r} = \{4, 3, 2, 1\}) = 1$ will be the same mutual information as
when $r = \{1, 2, 3, 4\}$ and $p(\hat{r} = \{1, 2, 3, 4\}) = 1$ and in
both of these cases the mutual information will be maximized at:

@@

# Weighted Mutual Information

In order to alleviate this problem, we instead consider *weighted* mutual information (@Guiasu1977-nu). We want to give more weight to permutations from $\hat{r}$ that are "closer" to $r$.  Essentially, this problem comes down to choosing a metric that measures the distance between the vectors $\hat{r}$ and $r$. One then gives more weight to combinations of $\hat{r}$ and $r$ that are closer and less weight to combinations that are far apart.  A natural choice for distance between vectors the sum of squared deviations between the two vectors.  Specifically, we define the loss function between $\hat{r}$ and $r$ to be $l(r, \hat{r}) = \sum_{i=1}^n (\hat{r}_i - r_i)^2$.  Then, since small losses should be associated with large weights, we focus on the recipricol of this loss function.

In general, any loss function $l(r, \hat{r})$ can be used to define the weighting function as follows:

@@

Note that in order to avoid dividing by 0, we define the weight to be 1 when $r$ and  $\hat{r}$ are exactly equal.

Specifically, when squared error loss is chosen, one gets the following weight function:

$$
w(r, \hat{r}) =
\begin{cases}
\frac{1}{\sum_{i=1}^n (\hat{r}_i - r_i)^2}, & r \ne \hat{r} \\
1, & r = \hat{r}
\end{cases}
$$ <!-- Note that one is 1/N! -->

@@

An alternative to defining weight this way is to consider looking at the correlation between $\hat{r}$ and $r$.  Since, $\hat{r}$ and $r$ are already ranks, the Pearson and Spearman correlation coefficients will be computationally exactly the same in this setting if that was the measure that someone chose to use.

Spearman correlation can be written as follows:

$$
\rho_{s} =  1 - \frac{6  \sum_{i=1}^n (\hat{r}_i - r_i)^2}{n(n^2 - 1)}
$$,

which means that Spearman correlation is simply (one minus) a scaling of squared error loss. Since $\rho_{s} \in [-1,1]$, but we desire weights $w(\hat{r},r) \in [0,1]$, we define $w(\hat{r},r) = \frac{(\rho_s + 1)}{2}$ guaranteeing weights in the desired range.

While squared error loss and Spearman/Pearson correlation work well, in this setting we argue that Kendall's $\tau$ correlation is the most appropriate for weighting due to its interpretation and how this related to competitions.  One way of interpreting Kendall's $\tau$ is that it's a scaling of the number of adjacent swaps required to go from one permutation of ranks to another. 
Recall that Kendall's $\tau$ is defined as: 
$$
\rho_\tau = 1 - \frac{2 \times \# \text{ discordant pairs}}{{n \choose 2}}
$$.
To convert $\rho_\tau$ back to the range between $[0,1]$, which we want for weighting purposes, we subtract 1 and divide by 2 giving us $\frac{\rho_\tau + 1}{2}$.  This then becomes:

$$
w(\hat{r},r) = \frac{\rho_\tau + 1}{2} = 1 - \frac{\# \text{ discordant pairs}}{{n \choose 2}}
$$ 
and the number of discordant pairs is equivalent to the number of adjacent swaps needed to get the order of $\hat{r}$ correct.

As an example, consider $r = (1,2,3,4)$ and $\hat{r} = (1,3,4,2)$.  This requires two adjacent swaps to get $\hat{r}$ to equal $r$ (i.e swap 4 with 2 and then 3 with 2). The Kendall's $\tau$ value between these two vectors is `r cor(c(1,2,3,4),c(1,3,4,2), method = "kendall")`.  We would then define the weight between the two vectors to be $(\rho_{\tau} + 1)/2 =$ `r (cor(c(1,2,3,4),c(1,3,4,2), method = "kendall") + 1)/2`.  This is exactly equivalent to $1 - \frac{\#\text{ adj swaps}}{{n \choose 2}}$.

<!-- ```{r} -->
<!-- cor(c(1,2,3,4),c(1,3,4,2), method = "kendall") -->
<!-- 1 - (2*2)/choose(4, 2) -->

<!-- cor(c(1,2,3,4),c(1,2,4,3), method = "kendall") -->
<!-- 1 - (2*1)/choose(4, 2) -->

<!-- #Converting this back to 0,1 gives (by adding 1 and dividing by 2):  -->
<!-- #(Number of swaps)/choose(4, 2).   -->
<!-- #This is what we use for weights -->
<!-- ``` -->

One and appropriate weight function was chosen, we get the following formula for weighted information:

$$
wI(r,\hat{r})= \sum_{r}\sum_{\hat{r}}w(r, \hat{r})p(r,\hat{r})\log \frac {p(r,\hat{r})}{p(r)p(\hat{r})}
$$

# Salient Weights (and ranks)

@shieh1998weighted: Weighted Kendall's Tau

Regardless of which method one chooses for defining weights between $\hat{r}$ and $r$, a second set of weights needs to be defined because the number of teams that need to be accurately ordered differs based on the choice of salient ranks. Therefore, we define a second set of weights, $m$, which is used to indicate which ranks are salient. We call these the "salient weights" to differentiate from the other weighting function, $w$. As an example, for a four-team tournament, if the organizer decided that only the correctly choosing the best team as the winner was important, then $m = \{1,0,0,0\}$ (i.e. essentially measuring effectivity)^[7].

^[7]: One can imagine that in an event like the Olympics, the salient weights would be 1's for the first three elements and 0's elsewhere.

In addition, one can also weight the salient ranks.  For example, in the Olmypics where getting the top three correct is the main objective, one could argue that it is more important to get the first place correct than the third place. This could lead to, for instance, a salient weight vector of $m = \{5,3,1,0, \cdots, 0\}$ giving 5 times the weight to getting the first place team correct as getting the third team correct.

With the addition of the salient weighting function, $m$, the formula for weighted mutual information becomes:

$$
wI_m(r_{\{{i|m_i >0}\}}, \hat{r}_{\{{i|m_i >0}\}})=\sum_{r_{\{{i|m_i >0}\}}}\sum_{\hat{r}_{\{{i|m_i >0}\}}} w_m(\hat{r}_{\{{i|m_i >0}\}}, r_{\{{i|m_i >0}\}}) \times
$$

$$
p(\hat{r}_{\{{i|m_i >0}\}}, r_{\{{i|m_i >0}\}})\times \\

\log \frac {p(\hat{r}_{\{{i|m_i >0}\}}, r_{\{{i|m_i >0}\}})}{p(r_{\{{i  \mid m_i >0}\}})p(\hat{r}_{\{{i \mid m_i >0}\}})}
$$

>>>>

However, using mutual information in this form does not suit our needs
in this setting. The problem is that mutual information in this form
will yield high values of mutual information when the output from the
tournament $\hat{r}$ is highly consistent *even if the ranking from the
tournament is incorrect*. As an example, if $n=4$ and the true order of
$\boldsymbol{\theta}$ is $r = \{1, 2, 3, 4\}$ and
$p(\hat{r} = \{4, 3, 2, 1\}) = 1$ will be the same mutual information as
when $r = \{1, 2, 3, 4\}$ and $p(\hat{r} = \{1, 2, 3, 4\}) = 1$ and in
both of these cases the mutual information will be maximized at:

@@

In order to alleviate this problem, we instead consider *weighted*
mutual information (@Guiasu1977-nu). We want to give more weight to
permutations from $\hat{r}$ that are "closer" to $r$.

In general, any loss function $l(r, \hat{r})$ can be used to define the
weigting function:

@@

Here though, we choose to use squared error loss:

$$
w(r, \hat{r}) =
\begin{cases}
\frac{1}{\hat{r}'r}, & r \ne \hat{r} \\
1, & r = \hat{r}
\end{cases}
$$ <!-- Note that one is 1/N! -->

@@

However, because the number of teams that need to be accurately ordered
differs based on the choice of salient ranks, a second set of weights is
defined, $m$, is used to indicate which ranks are salient. We call these
the "salient weights" to differentiate from the other weighting
function. As an example, for a four-team tournament, if the organizer
decides that only the winner is important, then $m = \{1,0,0,0\}$ (i.e.
essentially measuring effectivity). While here we only consider
indicators (i.e. 0's or 1's) in the salient weights vector, there is no
reason why you couldn't weight the ranks by relative levels of
importance. For instance, letting $m = \{3,1,0,0\}$ indicates that the
salient ranks are 1 and 2, but getting the true best competitor as the
winner is more important than ranking the true second best competitor
second.

With the addition of the weighting functions, the formula for mutual
information becomes:

$$
wI_m(r, \hat{r})=\sum_{r}\sum_{\hat{r}} w(r,\hat{r}) * m * p(r,\hat{r})\log \frac {p(r,\hat{r})}{p(r)p(\hat{r})}
$$
